A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2013-04-21 docs/paper_lncs.tex
Fix citations

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{float}
% \usepackage{caption}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage{url}
\usepackage{cite}
\urldef{\mailsa}\path|{rcoh,  georgiou, prazavi}@mit.edu|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Massachusetts Institute of Technology\\
77 Massachusetts Avenue  Cambridge, MA 02139, USA\\
\mailsa\\
\url{http://www.mit.edu}}

@@

\section{Introduction}
In this project we aim to analyze and improve the performance of Python's
dictionary or dict by changing its underlying implementation.  Python \cite{python} is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
several different data types as keys. However as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we predominantly focus on the study of new implementations of
hashes for the string type for the rest of the paper.

@@

On the other hand this hashing strategy has the tendency to fill contiguous
slots of the hash table. CPython tries to solve this behavior by utilizing a
good collision resolution strategy. The collision resolution strategy currently
used is open addressing with a custom non-linear probing scheme. In this scheme
CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
since again it behaves the same as linear-probing for some keys since it scans
the table entries in a fixed order. To overcome this issue CPython also uses the
most significant bits of the hash value with the following code snippet:

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing the tabulation tables in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
to generate these instructions.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

Our primary benchmark, hash-table-shootout, is a synthetic one that called the CPython dictionary
code directly from C so that all the interpreter overhead was bypassed.  We
measured performance for both sequential strings, generated by converting
sequential integers into their string representation, and for random strings,
generated by the string representation of random integers.  For each of the
benchmarks, we inserted a given number of keys into the dictionary and measured
the runtime.

@@

\subsection{Benchmarks}
Due to the heavy utilization of dictionaries in Python internals, one of the
valuable benchmarks is to see how the new dictionary performs with them them. We used
PyBench benchmark suite for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

As mentioned before, another set of the benchmarks \cite{hashshoot} that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

@@

\bibliography{references}{}
\bibliographystyle{plain}
\end{document}

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{float}
% \usepackage{caption}
\usepackage{setspace}
% \usepackage{fullpage}
\usepackage{url}
\usepackage{cite}
\urldef{\mailsa}\path|{rcoh,  georgiou, prazavi}@mit.edu|
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Massachusetts Institute of Technology\\
77 Massachusetts Avenue  Cambridge, MA 02139, USA\\
\mailsa}

@@

\section{Introduction}
In this paper we aim to analyze and improve the performance of Python's
dictionary or dict by changing its underlying implementation.  Python \cite{python} is an extremely popular general purpose computing language for several reasons:
\begin{enumerate}
 \item Clean, elegant near-pseudo code syntax leads to clear code 
 \item Widely distributed and available
 \item Large quantity of external libraries for specific features 
\end{enumerate}

@@

In this section we give a brief overview of the key implementation details of
dictionaries in CPython 2.7.3 \cite{python27}.  We should note that Python's dictionary supports
several different data types as keys. However, as we noted earlier, the
dictionaries underlying class instances and modules have only strings as keys.
Therefore optimizing a dictionary which have string-only keys is crucial to the
runtime of any Python program. CPython accommodates for this optimization by
changing its dictionary lookup method as necessary. First when a dictionary is
instantiated, CPython uses a string-only lookup function until a search for
non-string is requested, in this case CPython falls back to a more general
lookup method and uses this general function onward. There are two main
optimizations in the string-only lookup function.  First, since the string
comparisons do not raise any exceptions by design, some of the error checking
logic can been removed. Second, the string-only lookup function can skip the
rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
types can provide since string type does not have these special cases.  As it
can be seen, string-only keys can dominate the runtime of a dictionary and
CPython uses some measures to optimize for this common case. Because of this
importance we predominantly focus on the study of new implementations of
hashes for the string type for the rest of the paper.

@@

On the other hand, this hashing strategy has the tendency to fill contiguous
slots of the hash table. CPython tries to solve this behavior by utilizing a
good collision resolution strategy. The collision resolution strategy currently
used is open addressing with a custom non-linear probing scheme. In this scheme
CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
since again it behaves the same as linear-probing for some keys since it scans
the table entries in a fixed order. To overcome this issue CPython also uses the
most significant bits of the hash value with the following code snippet:

@@

After implementation and profiling, we found that dictionary table lookups had
over twice as many level 1 cache misses with tabulation hashing than without.
Hypothesizing that the tabulation tables were evicting other, more relevant
data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
instructions into our hash code to tell the CPU's caching system that it should
deprioritize storing the tabulation tables in cache.  We used GCC's
\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt} \cite{prefetch}
to generate these instructions.  However, this only yielded about a 1\% performance improvement,
negligible compared to the performance improvement of different hash schemes.

@@

Our primary benchmark, hash-table-shootout \cite{hashshoot}, is a synthetic one that called the CPython dictionary
code directly from C so that all the interpreter overhead was bypassed.  We
measured performance for both sequential strings, generated by converting
sequential integers into their string representation, and for random strings,
generated by the string representation of random integers.  For each of the
benchmarks, we inserted a given number of keys into the dictionary and measured
the runtime.

@@

\subsection{Benchmarks}
Due to the heavy utilization of dictionaries in Python internals, one of the
valuable benchmarks is to see how the new dictionary performs with them them. We used
PyBench benchmark suite \cite{pybench} for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on different standard real-world Python applications. We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite. The performance of more than ten applications were recoded as shown in the figures.

@@

\bibliography{references}{}
\bibliographystyle{plain}
